---
title: "HW1"
author: "Liancheng Lu"
date: "2024-01-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
setwd("/home/chenggg/BIOSTAT620")
Sys.setlocale("LC_TIME", "en_US.UTF-8")
rm(list = ls())
gc()
library(readxl)
library(dplyr)
library(ggplot2)
library(GGally)
library(lubridate)
library(reshape2)
library(tidyr)
df <- read_excel("ScreenTime.xlsx")
df <- df[c(1:11), ]
df$Pickup.1st_EST <- format(as.POSIXct(df$Pickup.1st_PST, format = "%H:%M", tz = "America/Los_Angeles"), "%H:%M", tz = "America/New_York")
```

# PATT I: DATA COLLECTION AND DATA PROCESSING

## Problem 1: Explore the your own screen activity data that you collect until the end of Friday (inclusive), January 26, 2024. This type of `break' time set up by scientists in practice is often referred to as data freeze date during data collection. If you were unable to collect such data due to the previous setting of your mobile device or other logistic reasons, please let the instructor or GSI know immediately, some backup data would be provided to you.

### a. Describe the purpose of the data collection, in which you state a scientific hypothesis of interest to justify your effort of data collection. Cite at least one reference to support your proposed hypothesis to be investigated. This hypothesis may be the one of a few possible hypotheses that you like to investigate in your first group project with your teammates.

### b. Explain the role of Informed Consent Form in connection to the planned study and data collection.

### c. Describe the data collection plan, including when the data is collected, which types of variables in the data are collected, where the data is collected from, and how many data are collected before the data freeze. You may use tables to summarize your answers if necessary.

### d. Create and add two new variables into your dataset; they are, "daily proportion of social screen time" (defined as the ratio of daily total social screen time over daily total screen time) and "daily duration per use" (defined as the ratio of daily total screen time over daily total of pickups).

```{r}
convert_to_minutes <- function(time) {
  if (!grepl("h", time)) {
    return(as.numeric(sub("m", "", time)))
  }
  parts <- strsplit(time, "h|m")[[1]]
  as.numeric(parts[1]) * 60 + as.numeric(parts[2])
}
df$Total.ST.min <- sapply(df$Total.ST, convert_to_minutes)
df$Social.ST.min <- sapply(df$Social.ST, convert_to_minutes)
df$prop_ST <- df$Social.ST.min / df$Total.ST.min
df$duration_per_use <- df$Total.ST.min / df$Pickups
head(df[,9:10])
```

## Problem 2: Data visualization is one of the early steps taken to see the data at hand. Consider the variables measured in the screen activity data, including daily total screen time, daily total social screen time, and daily number of pickups as well as two new variables derived from the raw data, daily proportion of social screen time and daily duration per use.

### a. Make a time series plot of each of the five variables in your data. Describe temporal patterns from these time series plots.

```{r}
# Daily total screen time
ggplot(data = df, aes(x = Date, y = Total.ST.min)) +
  geom_line() + 
  geom_point() +  
  labs(title = "Daily total screen time",
       x = "Date",
       y = "Minutes") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) 

# Daily total social screen time
ggplot(data = df, aes(x = Date, y = Social.ST.min)) +
  geom_line() + 
  geom_point() +  
  labs(title = "Daily total social screen time",
       x = "Date",
       y = "Minutes") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) 

# Daily number of pickups
ggplot(data = df, aes(x = Date, y = Pickups)) +
  geom_line() + 
  geom_point() +  
  labs(title = "Daily number of pickups",
       x = "Date",
       y = "Number of pickups") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) 

# Daily proportion of social screen time
ggplot(data = df, aes(x = Date, y = prop_ST)) +
  geom_line() + 
  geom_point() +  
  labs(title = "Daily proportion of social screen time",
       x = "Date",
       y = "Poportion of social screen time") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) 

# Daily duration per use
ggplot(data = df, aes(x = Date, y = duration_per_use)) +
  geom_line() + 
  geom_point() +  
  labs(title = "Daily duration per use",
       x = "Date",
       y = "Duration per use") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) 
```

### b. Make pairwise scatterplots of five variables. Describe correlation patterns from these pairwise scatterplots. Which pair of variables among the five variables has the highest correlation?

```{r}
df2 <- df[,c(1,3,5,6,9,10)]
ggpairs(df2[,-1]) 

df2_long <- pivot_longer(df2, cols = -Date, names_to = "series", values_to = "occupation_time")
ggplot(data = df2_long, aes(x = Date, y = occupation_time, color = series)) +
  geom_line() +
  labs(title = "Melted plot", x = "Time", y = "Count") +
  theme_minimal()
```

### c. Make an occupation time curve for each of the five time series. Explain the pattern of individual curves.

```{r}
# Calculate the P(x >= c) for Total.ST.min
df2 <- df2[order(df2$Total.ST.min, decreasing = TRUE),]  # Sorting by Total.ST.min
df2$Total.ST.prob <- seq_along(df2$Total.ST.min) / nrow(df2)

# Plot
ggplot(df2, aes(x = Total.ST.min, y = Total.ST.prob)) +
  geom_point() + 
  labs(x = "Vector Magnitude: c", 
       y = "P(x >= c)",
       title = "Occupation-Time Curves for total screen time") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) 

# Calculate the P(x >= c) for Social.ST.min
df2 <- df2[order(df2$Social.ST.min, decreasing = TRUE),]  # Sorting by Social.ST.min
df2$Social.ST.prob <- seq_along(df2$Social.ST.min) / nrow(df)

# Plot
ggplot(df2, aes(x = Social.ST.min, y = Social.ST.prob)) +
  geom_point() + 
  labs(x = "Vector Magnitude: c", 
       y = "P(x >= c)",
       title = "Occupation-Time Curves for social screen time") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) 

# Calculate the P(x >= c) for Pickups
df2 <- df2[order(df2$Pickups, decreasing = TRUE),]  # Sorting by Pickups
df2$Pickups.prob <- seq_along(df2$Pickups) / nrow(df)

# Plot
ggplot(df2, aes(x = Pickups, y = Pickups.prob)) +
  geom_point() + 
  labs(x = "Vector Magnitude: c", 
       y = "P(x >= c)",
       title = "Occupation-Time Curves for Number of Pickups") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) 

# Calculate the P(x >= c) for prop_ST
df2 <- df2[order(df2$prop_ST, decreasing = TRUE),]  # Sorting by prop_ST
df2$prop_ST.prob <- seq_along(df2$prop_ST) / nrow(df)

# Plot
ggplot(df2, aes(x = prop_ST, y = prop_ST.prob)) +
  geom_point() + 
  labs(x = "Vector Magnitude: c", 
       y = "P(x >= c)",
       title = "Occupation-Time Curves for Number of proportion of social screen time") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))

# Calculate the P(x >= c) for prop_ST
df2 <- df2[order(df2$duration_per_use, decreasing = TRUE),]  # Sorting by duration_per_use
df2$duration_per_use.prob <- seq_along(df2$duration_per_use) / nrow(df)

# Plot
ggplot(df2, aes(x = duration_per_use, y = duration_per_use.prob)) +
  geom_point() + 
  labs(x = "Vector Magnitude: c", 
       y = "P(x >= c)",
       title = "Occupation-Time Curves for Duration per use") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5))
```

### d. Use the R function acf to display the serial dependence for each of the five time series. Are there any significant autocorrelations? Explain your results. Note that in this R function, you may set plot=FALSE to yield values of the autocorrelations.

```{r fig.height=6}
par(mfrow=c(3,2))
for(i in 2:6) {
  acf_result <- acf(df2[[i]], main = paste("ACF for series", names(df2)[i]), plot = T)
  print(paste("Autocorrelations for series:", names(df2)[i]))
  print(acf_result$acf)
  acf_result
}
```

## Problem 3: Explore the use of the R package circular to display the time of first pickup as a circular variable or angular variable.

### a. Transform (or covert) the time of first pickup to an angle ranged from 0 to 360 degree, treating midnight as 0 degree. For example, 6AM is 90 degree and noon is 180 degree.

```{r}

```

